---
title: "Diabetes Binaries EDA and Predictive Model"
author: "jovi"
date: "2023-05-11"
output: html_document
---

```{r}
library(ggplot2)
library(readr)
library(dplyr)
```

```{r}
df <- read.csv("C:\\Users\\joviy\\diabetes_binary_health_indicators_BRFSS2015.csv")
```


```{r}
head(df)
```

```{r}
df <- subset(df, !duplicated(df))
summary(df)
```
Removing duplicates and summarize data


```{r}
colSums(is.na(df))
```
No missing Values

```{r}
dim(df)
```

```{r}
col_classes <- sapply(df, class)
col_classes
```
Print Column names
```{r}
hist(df$BMI, xlab = "BMI",
     col = "red", border = "black",
     xlim = c(0, 60),
     main = "BMI Frequency")
```

```{r}
hist(df$Age, xlab = "Age",
     col = "red", border = "black",
     main = "Age Frequency")
```

```{r}
ggplot(df, aes(x = factor(Diabetes_binary))) +
  geom_bar() +
  labs(title = "Count Plot", x = "Diabetes", y = "Count")
```
```{r}
hist(df$MentHlth, xlab = "Mental Health",
     col = "red", border = "black",
     main = "Mental Health Frequency")
```
```{r}
hist(df$MentHlth, xlab = "Education Level",
     col = "red", border = "black",
     main = "Education Frequency")
```


```{r}
ggplot(df, aes(x= factor(Diabetes_binary)),
       xlab="Diabetes Indicator") + geom_bar()
```

```{r}
# Create a new data frame with counts of each combination of diabetes_binary and highChol
counts <- df %>% group_by(Diabetes_binary, HighChol) %>% summarise(count = n())

# Create a stacked bar chart
ggplot(counts, aes(x = Diabetes_binary, y = count, fill = factor(HighChol))) +
  geom_bar(stat = "identity") +
  guides(fill = guide_legend(title = "HighChol"))
```

```{r}
# Create a new data frame with counts of each combination of diabetes_binary and Stroke
counts <- df %>% group_by(Diabetes_binary, Stroke) %>% summarise(count = n())

# Create a stacked bar chart
ggplot(counts, aes(x = Diabetes_binary, y = count, fill = factor(Stroke))) +
  geom_bar(stat = "identity") +
  guides(fill = guide_legend(title = "Stroke"))
```

```{r}
# Create a new data frame with counts of each combination of diabetes_binary and Smoker
counts <- df %>% group_by(Diabetes_binary, Smoker) %>% summarise(count = n())

# Create a stacked bar chart
ggplot(counts, aes(x = Diabetes_binary, y = count, fill = factor(Smoker))) +
  geom_bar(stat = "identity") +
  guides(fill = guide_legend(title = "Smoker"))
```

```{r}
# Create a new data frame with counts of each combination of diabetes_binary and HeartDiseaseorAttack
counts <- df %>% group_by(Diabetes_binary, HeartDiseaseorAttack) %>% summarise(count = n())

# Create a stacked bar chart
ggplot(counts, aes(x = Diabetes_binary, y = count, fill = factor(HeartDiseaseorAttack))) +
  geom_bar(stat = "identity") +
  guides(fill = guide_legend(title = "HeartDiseaseorAttack"))
```

STATISTICAL TESTING
```{r}
# Chi square test 

features <- df[,-1]

for (col in names(features)) {
  # Create a contingency table of the feature and the label
  table <- table(features[[col]], df$Diabetes_binary)
  
  # Perform the chi-square test
  result <- chisq.test(table)
  
  # Print the result
  cat("Chi-square test for", col, "\n")
  print(result)
  cat("\n")
}

```
in CHi2 test all variable has the p-value of <2.2e-16 which means the variables are significant
```{r}
# T-test

features <- df[,-1]

for (col in names(features)) {
  # Create a contingency table of the feature and the label
  table <- table(features[[col]], df$Diabetes_binary)
  
  # Perform the chi-square test
  result <- t.test(table)
  
  # Print the result
  cat("T- Test for", col, "\n")
  print(result)
  cat("\n")
}
```
Summary: using p-value < 0.05 for significancy from the test we get these variable that reject null hypothesis : BMI, GenHlth, Age,  Education, Income.
evaluting the scores of the variable BMI and Age is the top 2 highest t-test score meaning this 2 variable may contribute the most in predictive model

```{r}
# Anova Testing
fit <- lm(GenHlth ~ Education, data = df)
anova(fit)

features <- df[,-1]

for (col in names(features)) {
  # Perform the ANOVA test
  model <- aov(features[[col]] ~ df$Diabetes_binary)
  
  # Print the ANOVA table
  cat("ANOVA for", col, "\n")
  print(summary(model))
  cat("\n")
}
```
Summary :
in the test each of features have p-value of <2e-16 which means using anova all features has strong evidence to reject null hypotesis a high F-score means that it represent ratio of variance between group so in this test features with high scores are : HighBP, HighChol, BMI, HeartDiseaseorAttack, GenHlth , DiffWalk , Age , PhysHlth and Income may have significant contribution in the predictive model


```{r}
correlations <- cor(df)
target_correlations <- correlations["Diabetes_binary", ]
target_correlations

correlation_data <- melt(correlations)

ggplot(correlation_data, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "blue", high = "red") +
  theme_minimal() +
  labs(title = "Correlation Heatmap") +
  coord_equal(ratio = 1) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
  
```
Cheking correlations of features to the label

```{r}
library(rpart)
library(rpart.plot)

model <- rpart(Diabetes_binary ~ ., data = df)
```

```{r}
summary(model)
```


```{r}
library(caret)
varImp(model)
```
shows the variable importance meassure
```{r}
rpart.plot(model, extra = "auto")
```

```{r}
predictions <- predict(model, newdata = df)

# Convert predicted probabilities to class labels (., 0 or 1)
predicted_classes <- ifelse(predictions >= 0.5, 1, 0)

# Calculate accuracy
accuracy <- mean(predicted_classes == df$Diabetes_binary)

# Calculate ROC curve and AUC
library(pROC)
roc_obj <- roc(df$Diabetes_binary, predictions)
roc_auc <- auc(roc_obj)


# Print the accuracy and AUC
print(paste("Accuracy:", accuracy))
print(paste("ROC AUC:", roc_auc))
```
