---
title: "Diabetes Binaries EDA and Predictive Model"
author: "jovi"
date: "2023-05-11"
output: html_document
---

```{r}
library(ggplot2)
library(readr)
library(dplyr)
```

```{r}
df <- read.csv("C:\\Users\\joviy\\diabetes_binary_health_indicators_BRFSS2015.csv")
```


```{r}
head(df)
```

```{r}
df <- subset(df, !duplicated(df))
summary(df)
```
Removing duplicates and summarize data


```{r}
colSums(is.na(df))
```
No missing Values

```{r}
dim(df)
```

```{r}
col_classes <- sapply(df, class)
col_classes
```
Print Column names
```{r}
hist(df$BMI, xlab = "BMI",
     col = "red", border = "black",
     xlim = c(0, 60),
     main = "BMI Frequency")
```

```{r}
hist(df$Age, xlab = "Age",
     col = "red", border = "black",
     main = "Age Frequency")
```

```{r}
ggplot(df, aes(x = factor(Diabetes_binary))) +
  geom_bar() +
  labs(title = "Count Plot", x = "Diabetes", y = "Count")
```
```{r}
hist(df$MentHlth, xlab = "Mental Health",
     col = "red", border = "black",
     main = "Mental Health Frequency")
```
```{r}
hist(df$MentHlth, xlab = "Education Level",
     col = "red", border = "black",
     main = "Education Frequency")
```


```{r}
ggplot(df, aes(x= factor(Diabetes_binary)),
       xlab="Diabetes Indicator") + geom_bar()
```

```{r}
# Create a new data frame with counts of each combination of diabetes_binary and highChol
counts <- df %>% group_by(Diabetes_binary, HighChol) %>% summarise(count = n())

# Create a stacked bar chart
ggplot(counts, aes(x = Diabetes_binary, y = count, fill = factor(HighChol))) +
  geom_bar(stat = "identity") +
  guides(fill = guide_legend(title = "HighChol"))
```

```{r}
# Create a new data frame with counts of each combination of diabetes_binary and Stroke
counts <- df %>% group_by(Diabetes_binary, Stroke) %>% summarise(count = n())

# Create a stacked bar chart
ggplot(counts, aes(x = Diabetes_binary, y = count, fill = factor(Stroke))) +
  geom_bar(stat = "identity") +
  guides(fill = guide_legend(title = "Stroke"))
```

```{r}
# Create a new data frame with counts of each combination of diabetes_binary and Smoker
counts <- df %>% group_by(Diabetes_binary, Smoker) %>% summarise(count = n())

# Create a stacked bar chart
ggplot(counts, aes(x = Diabetes_binary, y = count, fill = factor(Smoker))) +
  geom_bar(stat = "identity") +
  guides(fill = guide_legend(title = "Smoker"))
```

```{r}
# Create a new data frame with counts of each combination of diabetes_binary and HeartDiseaseorAttack
counts <- df %>% group_by(Diabetes_binary, HeartDiseaseorAttack) %>% summarise(count = n())

# Create a stacked bar chart
ggplot(counts, aes(x = Diabetes_binary, y = count, fill = factor(HeartDiseaseorAttack))) +
  geom_bar(stat = "identity") +
  guides(fill = guide_legend(title = "HeartDiseaseorAttack"))
```

STATISTICAL TESTING
```{r}

# Filter binary and continuous columns
binary_columns <- df %>%
  select_if(function(x) all(x %in% c(0, 1)))

continuous_columns <- df %>%
  select_if(function(x) is.numeric(x) && !all(x %in% c(0, 1)))

# Print the filtered columns
dependent <- binary_columns$Diabetes_binary
binary_columns <- binary_columns[, -which(names(binary_columns) == "Diabetes_binary")]

ordinal_columns <- subset(continuous_columns, select = c(Education, Income, Age, GenHlth))
continuous_columns <- subset(continuous_columns, select = -c(Education, Income, Age, GenHlth))

print(ordinal_columns)

print(binary_columns)
print(continuous_columns)
print(dependent)
```


```{r}
print("Binary - Binary:")
for (var in colnames(binary_columns)) {
  # Binary - Binary (Chi-square and Phi coefficient)
  chi_square <- chisq.test(dependent, binary_columns[[var]])$statistic
  
  print(paste("Column name:", var))
  print(paste("Chi-square:", chi_square))
  
  
}

phi_coefficients <- vector()

for (i in 1:ncol(binary_columns)) {
  column <- binary_columns[, i]
  # print(column)
  cross_table <- table(column, dependent)
  n <- sum(cross_table)
  # Calculate observed frequencies
  n11 <- as.numeric(cross_table[2, 2])
  n10 <- as.numeric(cross_table[2, 1])
  n01 <- as.numeric(cross_table[1, 2])
  n00 <- as.numeric(cross_table[1, 1])

  # Calculate expected frequencies
  n1_ <- n11 + n10
  n0_ <- n01 + n00
  n_1 <- n11 + n01
  n_0 <- n10 + n00

  # Calculate phi coefficient
  phi <- (n11 * n00 - n10 * n01) / sqrt(n1_ * n0_ * n_1 * n_0)

  # Store the phi coefficient
  phi_coefficients <- c(phi_coefficients, phi)
}

print("")
print("-----------------------------------------------------------")
print("")

# Print phi coefficients for each binary column
for (i in 1:length(phi_coefficients)) {
  cat("Phi coefficient for", colnames(binary_columns)[i], ":", phi_coefficients[i], "\n")
}
```
Phi coefficient indicates from range [-1 ~ +1] if positive then the 2 variables compared has strong association if 0 then there is no association, if negative then the weak association

```{r}
# Chi square test 


for (col in names(continuous_columns)) {
  # Create a contingency table of the feature and the label
  table <- table(continuous_columns[[col]], dependent)
  
  # Perform the chi-square test
  result <- chisq.test(table)
  # Print the result
  cat("Chi-square test for", col, "\n")
  print(result)
  cat("\n")
}


```
in CHi2 test all variable has the p-value of <2.2e-16 which means the variables are significant
```{r}

# for (col in names(continuous_columns)) {
  correlation_rank <- cor.test(dependent, ordinal_columns$Education, method = "spearman")
  correlation_rank$estimate
# }

```

```{r}
library(ltm)
print("Binary - Ordinal:")
for (var in colnames(ordinal_columns)) {

  correlation_result <- biserial.cor(ordinal_columns[[var]], dependent)

  rank_biserial <- correlation_result
  
  print(paste("Column name:", var))
  print(paste("rank biserial:", rank_biserial))
  
  print(rank_biserial)
}
```

```{r}
# Mann-Whitney U test

for (col in names(ordinal_columns)) {
  # Create a contingency table of the feature and the label
  table <- table(ordinal_columns[[col]], dependent)
  
  # Perform the chi-square test
  mwu_result <- wilcox.test(table)

  # Print the result
  # cat("T- Test for", col, "\n")
  print(mwu_result)
  # cat("\n")
}
```

```{r}
# independent non parametric t-test

for (col in names(ordinal_columns)) {
  # Create a contingency table of the feature and the label
  table <- table(ordinal_columns[[col]], dependent)
  
  # Perform the chi-square test
  ttest_result <- t.test(table)

  # Print the result
  # cat("T- Test for", col, "\n")
  print(ttest_result)
  # cat("\n")
}
```




```{r}
library(rpart)
library(rpart.plot)

model <- rpart(Diabetes_binary ~ ., data = df)
```

```{r}
summary(model)
```


```{r}
library(caret)
varImp(model)
```
shows the variable importance meassure
```{r}
rpart.plot(model, extra = "auto")
```

```{r}
predictions <- predict(model, newdata = df)

# Convert predicted probabilities to class labels (., 0 or 1)
predicted_classes <- ifelse(predictions >= 0.5, 1, 0)

# Calculate accuracy
accuracy <- mean(predicted_classes == df$Diabetes_binary)

# Calculate ROC curve and AUC
library(pROC)
roc_obj <- roc(df$Diabetes_binary, predictions)
roc_auc <- auc(roc_obj)


# Print the accuracy and AUC
print(paste("Accuracy:", accuracy))
print(paste("ROC AUC:", roc_auc))
```
